---
title: "Bioregion: Northern California Depth Zones Analysis"
author: "Bailey Jørgensen"
format: html
editor: source
---

```{r}
# load libraries
librarian::shelf(tidyverse, janitor, sf, terra, tmap, knitr, showtext, gt, ggtext, leaflet, here)
```

## Here is what the PMEP depth zones stand for, according to the metadata:
### NOTE: The substrate data records the whole name, rather than using the # code

- Value 0: Landward Zone
- Value 1: Core Zone (Estuary Overlap)
- Value 2: Core Zone (Shoreline to -30m), State Waters
- Value 3: Core Zone (Shoreline to -30m), Federal Waters
- Value 4: Seaward Zone (-30m to -100m), State Waters
- Value 5: Seaward Zone (-30m to -100m), Federal Waters
- Value 6: Deep Shelf or Channel (-100m to -200m), State Waters
- Value 7: Deep Shelf or Channel (-100m to -200m), Federal Waters
- Value 8: Outside PMEP Score (>-300m) or International Waters

## For the purposes of the PMEP depth zone analysis, I have decided to use the PMEP Zones that are recorded in the substrate data, rather than the biotic data. This is because it is expected for there to be 100% MPA coverage of the different substrate types, where as the biotic data may not contain whole coverage of MPA, given the nature of biota. 

## Read in substrate, biota and MPA data

```{r}
# Read in substrate and biotic data...commented out because I chose to load the RDS files directly into the environment - uncomment to render
rds.dir <- "/capstone/marinebiomaps/data/rds-files/"
substrate <- sf::st_read(file.path(rds.dir, "substrate.rds"))

biota <- readRDS(here("data", "biota.rds")) |> 
  st_transform(biota, crs = 4326)


# Read in MPA boundaries data
boundary.dir <- "/capstone/marinebiomaps/data/MPA_boundaries"
mpa_boundaries <- sf::st_read(file.path(boundary.dir, "California_Marine_Protected_Areas_[ds582].shp"))

# Clean and transform MPA boundaries data
mpas <- mpa_boundaries |> 
  clean_names() |> 
#  select("type", "shortname", "geometry") |> 
  st_transform(mpas, crs = 4326) |> 
  st_make_valid()

# Filter the substrate data to just the variables I'm interested in - KEEP GEO
depth_zones_sub_filtered <- substrate |>
  select("pmep_zone_detail", "Shape", "pmep_region", "area_ha", "shape_area")
```

# Rather than filtering by PMEP region, lets try and filter by North, Central, and Southern CA
```{r}
colnames(mpas)
```
```{r}
unique(mpas$study_regi)
```

## Definitions: 

*SCSR* = South Coast Study Region
*NCCSR* = North Central Coast Study Region
*CCSR* = Central Coast Study Region
*NCSR* = North Coast Study Region
*SFBSR* = San Francisco Bay Study Region

```{r}
# Transform CRS for this map, since leaflet requires WGS84
mpas <- st_transform(mpas, crs = 4326)

# Factorize study regions
unique_regions <- c("NCSR", "NCCSR", "SFBSR", "CCSR", "SCSR")
mpas$study_regi <- factor(mpas$study_regi, levels = unique_regions)

# Define custom color palette
custom_colors <- c("#C4A9C2", "#71D9B0", "#D8C397", "#6EA1C8", "#796655")
pal <- colorFactor(palette = custom_colors, domain = unique_regions)

# Map with color-coded polygons
leaflet() |> 
  addProviderTiles(providers$CartoDB.Positron) |> 
  setView(lng = -119.6989, lat = 34.4203, zoom = 8) |> 
  addMiniMap(toggleDisplay = TRUE, minimized = FALSE) |> 
  addPolygons(data = mpas,
              color = ~pal(study_regi), # Assign colors based on study_region
              weight = 3,
              popup = paste0("MPA Name:  ", mpas$shortname, "<br>",
                             "MPA Type:  ", mpas$type, "<br>",
                             "Study Region:  ", mpas$study_regi)) |> 
  addLegend(position = "topright",
            pal = pal,
            values = mpas$study_regi,
            title = "MPA Study Region")
```


```{r}
# Filter to North
north_sub <- depth_zones_sub_filtered |>
  filter(pmep_region == "Pacific Northwest")

# All norcal MPAs with the substrate data within them
north_sub_mpa <- st_intersection(mpas, north_sub)

# Simple, no geo
north_sub_mpa_simple <- north_sub_mpa |> 
  st_drop_geometry()
```

```{r}
north_sub_mpa$shortname
```
Well, fuck, realizing that there is some name repeating here, and not all of the MPAs I expect are here...why might that be?

Make a basic Leaflet map of just the Norcal MPAs...to not include the substrate data, I will have to do some filtering, so I can use just the MPA data, but with the NorCal MPAS included in the PMEP Norcal Bioregion

```{r}
# Filter MPAs data to just these ones - ADD MORE THAT ARE NORCAL MPAS - FIGURE OUT WHATS WRONG WITH PMEP DATA FILTERING
shortname_list <- c(
  "Pyramid Point",
  "Southwest Seal Rock",
  "Castle Rock",
  "False Klamath Rock",
  "Reading Rock",
  "Samoa",
  "Point St. George Reef Offshore"
)

mpas_norcal <- mpas |> 
  filter(shortname %in% shortname_list)
  

# Transform CRS for this map, since leaflet requires WGS84
mpas <- st_transform(mpas, crs = 4326)

# Map
leaflet() |>
  addProviderTiles(providers$CartoDB.Positron) |>
  setView(lng = -119.6989,
          lat = 34.4203,
          zoom = 8) |>
  addMiniMap(toggleDisplay = TRUE,
             minimized = FALSE) |>
  addPolygons(data = mpas,
              color = "#09EEC8",
              weight = 3,
              popup = paste0("MPA Name:  ", mpas$shortname, "<br>",
                            "MPA Type:  ", mpas$type))

```

