---
title: "Bioregion: Southern California Depth Zones Analysis"
author: "Bailey Jørgensen"
format: html
---

```{r}
# load libraries
librarian::shelf(tidyverse, janitor, sf, terra, tmap, knitr, showtext, gt, ggtext, leaflet)
```

## Here is what the PMEP depth zones stand for, according to the metadata:
### NOTE: The substrate data records the whole name, rather than using the # code

- Value 0: Landward Zone
- Value 1: Core Zone (Estuary Overlap)
- Value 2: Core Zone (Shoreline to -30m), State Waters
- Value 3: Core Zone (Shoreline to -30m), Federal Waters
- Value 4: Seaward Zone (-30m to -100m), State Waters
- Value 5: Seaward Zone (-30m to -100m), Federal Waters
- Value 6: Deep Shelf or Channel (-100m to -200m), State Waters
- Value 7: Deep Shelf or Channel (-100m to -200m), Federal Waters
- Value 8: Outside PMEP Score (>-300m) or International Waters

## For the purposes of the PMEP depth zone analysis, I have decided to use the PMEP Zones that are recorded in the substrate data, rather than the biotic data. This is because it is expected for there to be 100% MPA coverage of the different substrate types, where as the biotic data may not contain whole coverage of MPA, given the nature of biota. 

## Read in substrate, biota and MPA data

```{r}
# Read in substrate and biotic data...commented out because I chose to load the RDS files directly into the environment - uncomment to render
substrate <- readRDS(here("data", "substrate.rds"))
biota <- readRDS(here("data", "biota.rds")) 

# Read in MPA boundaries data
boundary.dir <- "/capstone/marinebiomaps/data/MPA_boundaries"
mpa_boundaries <- sf::st_read(file.path(boundary.dir, "California_Marine_Protected_Areas_[ds582].shp"))

# Clean and transform MPA boundaries data
mpas <- mpa_boundaries |> 
  clean_names() |> 
  select("type", "shortname", "geometry") |> 
  st_transform(mpas, crs = st_crs(substrate)) |> 
  st_make_valid()

# Filter the substrate data to just the variables I'm interested in - KEEP GEO
depth_zones_sub_filtered <- substrate |>
  select("pmep_zone_detail", "Shape", "pmep_region")
```

```{r}
# Create variables for SoCal analysis (these were calculated in the Statewide analysis QMD)
south_sub <- depth_zones_sub_filtered |>
  filter(pmep_region == "Southern California Bight")

south_sub_mpa <- st_intersection(mpas, south_sub)

south_sub_mpa_simple <- south_sub_mpa |> 
  st_drop_geometry()
```