---
title: "Central Coast Study Region Depth Zones Analysis"
author: "Bailey Jørgensen"
format: html
editor: source
---

```{r}
# load libraries
librarian::shelf(tidyverse, janitor, sf, terra, tmap, knitr, showtext, gt, ggtext, leaflet)
```

## Here is what the PMEP depth zones stand for, according to the metadata:

### NOTE: The substrate data records the whole name, rather than using the # code

-   Value 0: Landward Zone
-   Value 1: Core Zone (Estuary Overlap)
-   Value 2: Core Zone (Shoreline to -30m), State Waters
-   Value 3: Core Zone (Shoreline to -30m), Federal Waters
-   Value 4: Seaward Zone (-30m to -100m), State Waters
-   Value 5: Seaward Zone (-30m to -100m), Federal Waters
-   Value 6: Deep Shelf or Channel (-100m to -200m), State Waters
-   Value 7: Deep Shelf or Channel (-100m to -200m), Federal Waters
-   Value 8: Outside PMEP Score (\>-300m) or International Waters

## For the purposes of the PMEP depth zone analysis, I have decided to use the PMEP Zones that are recorded in the substrate data, rather than the biotic data. This is because it is expected for there to be 100% MPA coverage of the different substrate types, where as the biotic data may not contain whole coverage of MPA, given the nature of biota.

## Read in substrate, biota and MPA data

```{r}
# Read in substrate and biotic data
#substrate <- readRDS("substrate.rds")
#biota <- readRDS("Biota.rds") |>
#  clean_names()

biota <- Biota |>
  clean_names()

# Read in MPA boundaries data
boundary.dir <- "/capstone/marinebiomaps/data/MPA_boundaries"
mpa_boundaries <- sf::st_read(file.path(boundary.dir, "California_Marine_Protected_Areas_[ds582].shp"))
```

### Combine the MPA boundary data with the substrate data

This code chunk filters ALL mpas - keeps geo

```{r}
# Clean boundaries data and get it ready for joining - Michelle's code is commented out. Use?
mpas <- mpa_boundaries |> 
  clean_names() |> 
  select("type", "shortname", "geometry") |> 
  st_transform(mpas, crs = st_crs(substrate))
  
  #filter(name == "Point Lobos SMR") %>%
  # Select relevant columns
  # select(-c("objectid", "ccr", "area_sq_mi", "study_regi", "fullname", "ccr_int", "name", "acres", "shape_are", "shape_len")) %>%
  # mutate(area_km_mpa = hectares / 100) %>%  # Convert hectares to square km
  # select(-hectares)
```

This code chunk selects substrate cats i want - keeps geo (REVISIT THIS IS MORE CATEGORIES ARE NEEDED IN THE FUTURE)

```{r}
# FIrst, filter the substrate data to just the variables I'm interested in
depth_zones_sub_filtered <- substrate |>
  select("pmep_zone_detail", "Shape", "pmep_region")
```

This code chunk combines substrate and mpa...but just for Point Lobos

```{r}
# Filter mpa data to just Point Lobos
point_lobos <- mpas |>
  filter(shortname == "Point Lobos")

# Perform spatial join
lobos_substrate_depth_geo <- st_intersection(point_lobos, depth_zones_sub_filtered)
```

Rename the pmep zones

```{r}
lobos_substrate_depth_geo <- lobos_substrate_depth_geo %>%
  mutate(pmep_zone_detail = recode(pmep_zone_detail,
                                   "Core Zone (Shoreline to -30m), State Waters" = "Zone 2: Core Zone (Shoreline to -30m), State Waters",
                                   "Deep Shelf or Sound (-100m to -200m), State Waters" = "Zone 6: Deep Shelf or Sound (-100m to -200m), State Waters",
                                   "Landward Zone" = "Zone 0: Landward Zone",
                                   "Outside PMEP Scope (>-200m) or International Waters" = "Zone 8: Outside PMEP Scope (>-200m) or International Waters",
                                   "Seaward Zone (-30m to -100m), State Waters" = "Zone 4: Seaward Zone (-30m to -100m), State Waters"))  
```

```{r}
unique(lobos_substrate_depth_geo$pmep_zone_detail)
```

### This chunk makes a map of Point Lobos Depth Zones!

```{r}
font_add_google("Montserrat", "montserrat")
showtext_auto()

# Create base map with the colored polygons
tm_shape(lobos_substrate_depth_geo) +
  tm_polygons(col = "pmep_zone_detail",
              palette = c('#09EEC8','#6EA1C8', '#3C7BB7'),
              title = "Depth\nZones",
              labels = c("Zone 0", "Zone 2", "Zone 4", "Zone 6", "Zone 8")) +
  # Add the same shape again but with solid fill to mask underlying features
  tm_shape(lobos_substrate_depth_geo) +
  tm_fill(col = "pmep_zone_detail",
          palette = c('#09EEC8','#6EA1C8', '#3C7BB7'),
          legend.show = FALSE) +  # Don't show this in legend
  tm_layout(legend.position = c("right", "bottom"),
            legend.title.fontfamily = "montserrat",
            legend.text.fontfamily = "montserrat") 
  
```

# Calculate the percentage of each depth zone in the Point Lobos MPA

```{r}
# Drop geometry
lobos_substrate_depth_simple <- lobos_substrate_depth_geo |>
  st_drop_geometry()
```

```{r}
# Step 2: Make sure pmep_zone is treated as a factor (categorical)
lobos_substrate_depth_simple$pmep_zone_detail <- as.factor(lobos_substrate_depth_simple$pmep_zone_detail)

# Step 3: Calculate the percentages for each shortname and pmep_zone combination
lobos_percentages <- lobos_substrate_depth_simple |>
  # Group by shortname and pmep_zone to count occurrences
  group_by(shortname, pmep_zone_detail) |>
  summarise(count = n(), .groups = "drop") |>
  
  # Group by shortname to calculate total count for each shortname
  group_by(shortname) %>%
  mutate(percentage = count / sum(count) * 100) |>
  
  # Reshape data to have a column for each unique pmep_zone
  select(-count) |>
  pivot_wider(
    names_from = pmep_zone_detail,
    values_from = percentage) 

```

```{r}
# Make a lobos pivot longer table
lobos_long <- lobos_percentages %>%
  pivot_longer(
    cols = -`shortname`, 
    names_to = "PMEP_Zone",
    values_to = "Percentage"
  )
```

### Make a table presenting depth zones averages for Point Lobos!

```{r}

lobos_long |>
  gt() |> 
  tab_header(
    title = md("Depth Zones of **Point Lobos SMR and SMCA**"),
    subtitle = "depth zones as percentages of total protected area"
  ) |> 
  tab_style(
    style = cell_text(size = px(12)),
    locations = cells_column_labels()
  ) |>
  tab_style(
    style = cell_text(size = px(10)), # Adjust size for source note
    locations = cells_source_notes()
  ) |>
  fmt_number(
    columns = Percentage,  
    decimals = 2,
    suffix = "%"
  ) |>
  text_transform(
    locations = cells_body(columns = Percentage),
    fn = function(x) paste0(x, "%")  # Manually appends the % sign
  ) |>
  tab_style(
    style = cell_text(color = "#002f41"),  # Change text color (modify as needed)
    locations = cells_body(columns = Percentage)
  ) |> 
  tab_source_note(md("*Depth Zones classified by the Pacific Marine and Estuarine Partnership (PMEP)*")) 

```