---
title: "Central Coast Study Region Depth Zones Analysis"
author: "Bailey Jørgensen"
format: html
editor: source
---

```{r}
# load libraries
librarian::shelf(tidyverse, janitor, sf, terra, tmap, knitr, showtext, gt, ggtext, leaflet)
```

## Here is what the PMEP depth zones stand for, according to the metadata:

### NOTE: The substrate data records the whole name, rather than using the # code

-   Value 0: Landward Zone
-   Value 1: Core Zone (Estuary Overlap)
-   Value 2: Core Zone (Shoreline to -30m), State Waters
-   Value 3: Core Zone (Shoreline to -30m), Federal Waters
-   Value 4: Seaward Zone (-30m to -100m), State Waters
-   Value 5: Seaward Zone (-30m to -100m), Federal Waters
-   Value 6: Deep Shelf or Channel (-100m to -200m), State Waters
-   Value 7: Deep Shelf or Channel (-100m to -200m), Federal Waters
-   Value 8: Outside PMEP Score (\>-300m) or International Waters

## For the purposes of the PMEP depth zone analysis, I have decided to use the PMEP Zones that are recorded in the substrate data, rather than the biotic data. This is because it is expected for there to be near 100% MPA coverage of the different substrate types, where as the biotic data may not contain whole coverage of MPA, given the nature of biota.

## Read in substrate, biota and MPA data

```{r}
# Read in data
rds.dir <- "/capstone/marinebiomaps/data/rds-files"

# Read in MPA boundaries data
boundary.dir <- "/capstone/marinebiomaps/data/MPA_boundaries"
mpa_boundaries <- sf::st_read(file.path(boundary.dir, "California_Marine_Protected_Areas_[ds582].shp"))

# Clean and transform MPA boundaries data
mpas <- mpa_boundaries |> 
  clean_names() |> 
  st_transform(mpas, crs = st_crs(substrate)) |> 
  st_make_valid()

```

This code chunk combines substrate and mpa...but just for Point Lobos

```{r}
# Filter mpa data to just Point Lobos
point_lobos <- mpas |>
  filter(shortname == "Point Lobos")

# Perform spatial join
lobos_substrate_depth_geo <- st_intersection(point_lobos, depth_zones_sub_filtered)
```

Rename the pmep zones

```{r}
lobos_substrate_depth_geo <- lobos_substrate_depth_geo %>%
  mutate(pmep_zone_detail = recode(pmep_zone_detail,
                                   "Core Zone (Shoreline to -30m), State Waters" = "Zone 2: Core Zone (Shoreline to -30m), State Waters",
                                   "Deep Shelf or Sound (-100m to -200m), State Waters" = "Zone 6: Deep Shelf or Sound (-100m to -200m), State Waters",
                                   "Landward Zone" = "Zone 0: Landward Zone",
                                   "Outside PMEP Scope (>-200m) or International Waters" = "Zone 8: Outside PMEP Scope (>-200m) or International Waters",
                                   "Seaward Zone (-30m to -100m), State Waters" = "Zone 4: Seaward Zone (-30m to -100m), State Waters"))  
```

```{r}
unique(lobos_substrate_depth_geo$pmep_zone_detail)
```

### This chunk makes a map of Point Lobos Depth Zones!

```{r}
font_add_google("Montserrat", "montserrat")
showtext_auto()

# Create base map with the colored polygons
tm_shape(lobos_substrate_depth_geo) +
  tm_polygons(col = "pmep_zone_detail",
              palette = c('#09EEC8','#6EA1C8', '#3C7BB7'),
              title = "Depth\nZones",
              labels = c("Zone 0", "Zone 2", "Zone 4", "Zone 6", "Zone 8")) +
  # Add the same shape again but with solid fill to mask underlying features
  tm_shape(lobos_substrate_depth_geo) +
  tm_fill(col = "pmep_zone_detail",
          palette = c('#09EEC8','#6EA1C8', '#3C7BB7'),
          legend.show = FALSE) +  # Don't show this in legend
  tm_layout(legend.position = c("right", "bottom"),
            legend.title.fontfamily = "montserrat",
            legend.text.fontfamily = "montserrat") 
  
```

# Calculate the percentage of each depth zone in the Point Lobos MPA

```{r}
# Drop geometry
lobos_substrate_depth_simple <- lobos_substrate_depth_geo |>
  st_drop_geometry()
```

```{r}
# Step 2: Make sure pmep_zone is treated as a factor (categorical)
lobos_substrate_depth_simple$pmep_zone_detail <- as.factor(lobos_substrate_depth_simple$pmep_zone_detail)

# Step 3: Calculate the percentages for each shortname and pmep_zone combination
lobos_percentages <- lobos_substrate_depth_simple |>
  # Group by shortname and pmep_zone to count occurrences
  group_by(shortname, pmep_zone_detail) |>
  summarise(count = n(), .groups = "drop") |>
  
  # Group by shortname to calculate total count for each shortname
  group_by(shortname) %>%
  mutate(percentage = count / sum(count) * 100) |>
  
  # Reshape data to have a column for each unique pmep_zone
  select(-count) |>
  pivot_wider(
    names_from = pmep_zone_detail,
    values_from = percentage) 

```

```{r}
# Make a lobos pivot longer table
lobos_long <- lobos_percentages %>%
  pivot_longer(
    cols = -`shortname`, 
    names_to = "PMEP_Zone",
    values_to = "Percentage"
  )
```

### Make a table presenting depth zones averages for Point Lobos!

```{r}

lobos_long |>
  gt() |> 
  tab_header(
    title = md("Depth Zones of **Point Lobos SMR and SMCA**"),
    subtitle = "depth zones as percentages of total protected area"
  ) |> 
  tab_style(
    style = cell_text(size = px(12)),
    locations = cells_column_labels()
  ) |>
  tab_style(
    style = cell_text(size = px(10)), # Adjust size for source note
    locations = cells_source_notes()
  ) |>
  fmt_number(
    columns = Percentage,  
    decimals = 2,
    suffix = "%"
  ) |>
  text_transform(
    locations = cells_body(columns = Percentage),
    fn = function(x) paste0(x, "%")  # Manually appends the % sign
  ) |>
  tab_style(
    style = cell_text(color = "#002f41"),  # Change text color (modify as needed)
    locations = cells_body(columns = Percentage)
  ) |> 
  tab_source_note(md("*Depth Zones classified by the Pacific Marine and Estuarine Partnership (PMEP)*")) 

```