---
title: "North Coast Study Region Depth Zones Analysis"
author: "Bailey Jørgensen"
format: html
editor: source
---

```{r}
# load libraries
librarian::shelf(tidyverse, janitor, sf, terra, tmap, knitr, showtext, gt, ggtext, leaflet, here)
```

## Here is what the PMEP depth zones stand for, according to the metadata:
### NOTE: The substrate data records the whole name, rather than using the # code

- Value 0: Landward Zone
- Value 1: Core Zone (Estuary Overlap)
- Value 2: Core Zone (Shoreline to -30m), State Waters
- Value 3: Core Zone (Shoreline to -30m), Federal Waters
- Value 4: Seaward Zone (-30m to -100m), State Waters
- Value 5: Seaward Zone (-30m to -100m), Federal Waters
- Value 6: Deep Shelf or Channel (-100m to -200m), State Waters
- Value 7: Deep Shelf or Channel (-100m to -200m), Federal Waters
- Value 8: Outside PMEP Score (>-300m) or International Waters

## For the purposes of the PMEP depth zone analysis, I have decided to use the PMEP Zones that are recorded in the substrate data, rather than the biotic data. This is because it is expected for there to be near 100% MPA coverage of the different substrate types, where as the biotic data may not contain whole coverage of MPA, given the nature of biota. 

```{r}
# Read in ncsr substrate data
rds.dir <- "/capstone/marinebiomaps/data/rds-files/"
ncsr_substrate <- readRDS(file.path(rds.dir, "ncsr_substrate.rds"))

# Read in MPA boundaries data
boundary.dir <- "/capstone/marinebiomaps/data/MPA_boundaries"
mpa_boundaries <- sf::st_read(file.path(boundary.dir, "California_Marine_Protected_Areas_[ds582].shp"))

# Clean and transform MPA boundaries data
mpas <- mpa_boundaries |> 
  clean_names() |> 
#  select("type", "shortname", "geometry") |> 
  st_transform(mpas, crs = 4326) |> 
  st_make_valid()
```

