---
title: "North Coast Study Region Depth Zones Analysis"
author: "Bailey Jørgensen"
format: html
editor: source
---

```{r}
# load libraries
librarian::shelf(tidyverse, janitor, sf, terra, tmap, knitr, showtext, ggtext, leaflet, here)
```

## Here is what the PMEP depth zones stand for, according to the metadata:
### NOTE: The substrate data records the whole name, rather than using the # code

- Value 0: Landward Zone
- Value 1: Core Zone (Estuary Overlap)
- Value 2: Core Zone (Shoreline to -30m), State Waters
- Value 3: Core Zone (Shoreline to -30m), Federal Waters
- Value 4: Seaward Zone (-30m to -100m), State Waters
- Value 5: Seaward Zone (-30m to -100m), Federal Waters
- Value 6: Deep Shelf or Channel (-100m to -200m), State Waters
- Value 7: Deep Shelf or Channel (-100m to -200m), Federal Waters
- Value 8: Outside PMEP Score (>-300m) or International Waters

## For the purposes of the PMEP depth zone analysis, I have decided to use the PMEP Zones that are recorded in the substrate data, rather than the biotic data. This is because it is expected for there to be near 100% MPA coverage of the different substrate types, where as the biotic data may not contain whole coverage of MPA, given the nature of biota. 

```{r}
# Read in ncsr substrate and biota data
rds.dir <- "/capstone/marinebiomaps/data/rds-files/"
ncsr_substrate <- readRDS(file.path(rds.dir, "ncsr_substrate.rds"))
ncsr_substrate_fullregion <- readRDS(file.path(rds.dir, "ncsr_substrate_fullregion.rds"))
ncsr_biota <- readRDS(file.path(rds.dir, "ncsr_biota.rds"))
ncsr_biota_fullregion <- readRDS(file.path(rds.dir, "ncsr_biota_fullregion.rds"))
```

```{r}
# Make a simple dataframe
ncsr_substrate_simple <- ncsr_substrate |> 
  st_drop_geometry()

ncsr_substrate_fullregion_simple <- ncsr_substrate_fullregion |> 
  st_drop_geometry()

# Make a simple dataframe
ncsr_biota_simple <- ncsr_biota |> 
  st_drop_geometry()

ncsr_biota_fullregion_simple <- ncsr_biota_fullregion |> 
  st_drop_geometry()

```

```{r}
# Take a look at all the PMEP depth zones to verify
unique(ncsr_substrate_fullregion_simple$pmep_zone_detail)
```

```{r}
# Rename depth zone variables for searchability/ease - I will bring back more technical names for viz
ncsr_substrate_simple <- ncsr_substrate_simple |> 
  mutate(pmep_zone_detail = case_when(
    pmep_zone_detail == "Deep Shelf or Sound (-100m to -200m), State Waters"  ~ "Zone 6",
    pmep_zone_detail == "Outside PMEP Scope (>-200m) or International Waters" ~ "Zone 8",
    pmep_zone_detail == "Seaward Zone (-30m to -100m), Federal"               ~ "Zone 5",
    pmep_zone_detail == "Core Zone (Shoreline to -30m), State Waters"         ~ "Zone 2",
    pmep_zone_detail == "Landward Zone"                                       ~ "Zone 0",
    pmep_zone_detail == "Seaward Zone (-30m to -100m), State Waters"          ~ "Zone 4",
    pmep_zone_detail == "Deep Shelf or Sound (-100m to -200m), Federal"       ~ "Zone 7",
    pmep_zone_detail == "Core Zone (Shoreline to -30m), Federal"              ~ "Zone 3"
  ))

ncsr_substrate_fullregion_simple <- ncsr_substrate_fullregion_simple |> 
  mutate(pmep_zone_detail = case_when(
    pmep_zone_detail == "Deep Shelf or Sound (-100m to -200m), State Waters"  ~ "Zone 6",
    pmep_zone_detail == "Outside PMEP Scope (>-200m) or International Waters" ~ "Zone 8",
    pmep_zone_detail == "Seaward Zone (-30m to -100m), Federal"               ~ "Zone 5",
    pmep_zone_detail == "Core Zone (Shoreline to -30m), State Waters"         ~ "Zone 2",
    pmep_zone_detail == "Landward Zone"                                       ~ "Zone 0",
    pmep_zone_detail == "Seaward Zone (-30m to -100m), State Waters"          ~ "Zone 4",
    pmep_zone_detail == "Deep Shelf or Sound (-100m to -200m), Federal"       ~ "Zone 7",
    pmep_zone_detail == "Core Zone (Shoreline to -30m), Federal"              ~ "Zone 3"
  ))



unique(ncsr_substrate_fullregion_simple$pmep_zone_detail)
```

```{r}
head()
```


```{r}
# Total area across NCSR MPAs
total_ha_ncsr_mpas <- sum(ncsr_substrate_simple$area_ha, na.rm = TRUE)

print(paste0("The total area of Substrate data within NCSR MPAs in Ha: ", round(total_ha_ncsr_mpas, 2)))

# Summarize total area by zone and calculate percentage
zone_areas_ncsr <- ncsr_substrate_simple |> 
  group_by(pmep_zone_detail) |> 
  summarize(total_area_per_zone_ncsr = sum(area_ha, na.rm = TRUE), .groups = "drop") |> 
  mutate(percent_of_total_ncsr = (total_area_per_zone_ncsr / total_ha_ncsr_mpas) * 100)

# Print formatted output
zone_areas_ncsr |> 
  arrange(pmep_zone_detail) |> 
  rowwise() |> 
  mutate(print_statement = paste0(
    pmep_zone_detail, 
    ": ", round(total_area_per_zone_ncsr, 2), " ha (", 
    round(percent_of_total_ncsr, 2), "% of total)"
  )) |> 
  pull(print_statement) |> 
  cat(sep = "\n")

```

```{r}
# Total area across NCSR MPAs
total_ha_ncsr_fullregion <- sum(ncsr_substrate_fullregion_simple$area_ha, na.rm = TRUE)

print(paste0("The total area of Substrate data within NCSR Statewaters: ", round(total_ha_ncsr_fullregion, 3)))

# Summarize total area by zone and calculate percentage
zone_areas_ncsr_fullregion <- ncsr_substrate_fullregion_simple |> 
  group_by(pmep_zone_detail) |> 
  summarize(total_area_per_zone_ncsr_fullregion = sum(area_ha, na.rm = TRUE), .groups = "drop") |> 
  mutate(percent_of_total_ncsr_fullregion = (total_area_per_zone_ncsr_fullregion / total_ha_ncsr_fullregion) * 100)

# Print formatted output
zone_areas_ncsr_fullregion |> 
  arrange(pmep_zone_detail) |> 
  rowwise() |> 
  mutate(print_statement = paste0(
    pmep_zone_detail, 
    ": ", round(total_area_per_zone_ncsr_fullregion, 3), " ha (", 
    round(percent_of_total_ncsr_fullregion, 3), "% of total)"
  )) |> 
  pull(print_statement) |> 
  cat(sep = "\n")

```


**NOTE: the fact that the "federal waters" zones have anything at all in them is strange, considering that we have already filtered to MPAs, which I assume are all state waters? But maybe not? Bears clarifying.**

```{r}
# Taking out zone 1 since it is empty....as well as the federal zones
zones_to_plot <- c("Zone 0", "Zone 2", "Zone 4", "Zone 6", "Zone 8")

zone_areas_filtered_ncsr <- zone_areas_ncsr |> 
  filter(pmep_zone_detail %in% zones_to_plot) |> 
  mutate(pmep_zone_detail = case_when(
    pmep_zone_detail == "Zone 0" ~ "Landward Zone",
    pmep_zone_detail == "Zone 2" ~ "Core Zone (Shoreline to -30m)",
    pmep_zone_detail == "Zone 4" ~ "Seaward Zone (-30m to -100m)",
    pmep_zone_detail == "Zone 6" ~ "Deep Shelf or Sound (-100m to -200m)",
    pmep_zone_detail == "Zone 8" ~ "Outside PMEP Scope (>-200m)"
  )) |> 
  mutate(pmep_zone_detail = factor(pmep_zone_detail, levels = c(
    "Landward Zone",
    "Core Zone (Shoreline to -30m)",
    "Seaward Zone (-30m to -100m)",
    "Deep Shelf or Sound (-100m to -200m)",
    "Outside PMEP Scope (>-200m)"
  )))

font_add_google("Montserrat", "montserrat")
showtext_auto()

ggplot(zone_areas_filtered_ncsr, aes(x = percent_of_total_ncsr, y = pmep_zone_detail, fill = pmep_zone_detail)) +
  geom_col() +
  labs(
    title = "Depth Zones as a % of Total NCSR MPA Area",
    x = "Percent of Total NCSR Area (%)",
    y = "PMEP Depth Zone"
  ) +
  theme_classic() +
  theme(
    text = element_text(family = "montserrat", size = 18),
    axis.text.x = element_text(hjust = 1),
    legend.position = "none") +
  geom_label(aes(label = paste0(round(percent_of_total_ncsr, 1), "%")),
             fill = "white", 
             color = "black",
             hjust = 0,    # Align to left
             x = 1,        # Set near y-axis (adjust as needed)
             size = 3.5) +
  scale_fill_brewer(palette = "Blues") +
  scale_y_discrete(labels = function(x) str_wrap(x, width = 20))

```

```{r}
# Transform CRS for this map, since leaflet requires WGS84
mpas <- st_transform(mpas, crs = 4326)
ncsr_substrate <- st_transform(ncsr_substrate, crs = 4326)

#ncsr_substrate_simplify <- st_simplify(ncsr_substrate_simplify, dTolerance = 40000)
  

ncsr_substrate$cmecs_sc_category <- as.character(ncsr_substrate$cmecs_sc_category)

# Define custom color palette
substrate_categories <- unique(ncsr_substrate$cmecs_sc_category)
substrate_palette <- c("#204035FF", "#4A7169FF", "#849383", "#BEB59CFF", "#998467", "#735231FF", "#49271BFF")
pal <- colorFactor(palette = substrate_palette, domain = substrate_categories)

# Map with color-coded polygons
leaflet() |> 
  addProviderTiles(providers$CartoDB.Positron) |> 
  setView(lng = -124.3917, lat = 40.4667, zoom = 8) |> 
  addMiniMap(toggleDisplay = TRUE, minimized = FALSE) |> 
  addPolygons(data = mpas,
              color = "goldenrod", # Assign colors based on study_region
              weight = 2,
              fillOpacity = 0,     # No fill
              
              popup = paste0("MPA Name:  ", mpas$shortname, "<br>",
                             "MPA Type:  ", mpas$type, "<br>",
                             "Study Region:  ", mpas$study_regi)) |> 
  addPolygons(data = ncsr_substrate,
              color = ~pal(cmecs_sc_category), # Assign colors based on type
              weight = 3,
              fillOpacity = 1,
              popup = paste0("Substrate type:  ", ncsr_substrate$cmecs_sc_category, "<br>",
                              "Substrate type area:  ", ncsr_substrate$area_ha, "<br>")) |> 
  addLegend(position = "topright",
            pal = pal,
            values = ncsr_substrate$cmecs_sc_category,
            title = "Substrate Within MPAs")
```













