---
title: "California Statewide Depth Zones Analysis"
author: "Bailey Jørgensen"
format: html
editor: visual
---

```{r}
# load libraries
librarian::shelf(tidyverse, janitor, sf, terra, tmap, knitr, showtext, gt, ggtext, leaflet)
```

## Here is what the PMEP depth zones stand for, according to the metadata:
### NOTE: The substrate data records the whole name, rather than using the # code

- Value 0: Landward Zone
- Value 1: Core Zone (Estuary Overlap)
- Value 2: Core Zone (Shoreline to -30m), State Waters
- Value 3: Core Zone (Shoreline to -30m), Federal Waters
- Value 4: Seaward Zone (-30m to -100m), State Waters
- Value 5: Seaward Zone (-30m to -100m), Federal Waters
- Value 6: Deep Shelf or Channel (-100m to -200m), State Waters
- Value 7: Deep Shelf or Channel (-100m to -200m), Federal Waters
- Value 8: Outside PMEP Score (>-300m) or International Waters

## For the purposes of the PMEP depth zone analysis, I have decided to use the PMEP Zones that are recorded in the substrate data, rather than the biotic data. This is because it is expected for there to be 100% MPA coverage of the different substrate types, where as the biotic data may not contain whole coverage of MPA, given the nature of biota. 

## Read in substrate, biota and MPA data

```{r}
# Read in substrate and biotic data ... Commented out, because I chose to load the RDS files into my environment prior to this
#substrate <- readRDS("substrate.rds")
#biota <- readRDS("Biota.rds") |>
#  clean_names()

biota <- Biota |>
  clean_names()

# Read in MPA boundaries data
boundary.dir <- "/capstone/marinebiomaps/data/MPA_boundaries"
mpa_boundaries <- sf::st_read(file.path(boundary.dir, "California_Marine_Protected_Areas_[ds582].shp"))

# Clean and transform MPA boundaries data
mpas <- mpa_boundaries |> 
  clean_names() |> 
  select("type", "shortname", "geometry") |> 
  st_transform(mpas, crs = st_crs(substrate)) |> 
  st_make_valid()

# Filter the substrate data to just the variables I'm interested in - KEEP GEO
depth_zones_sub_filtered <- substrate |>
  select("pmep_zone_detail", "Shape", "pmep_region")
```

## OK HERE... I AM TRYING TO COMBINE THE DATA SO THAT I CAN DO FULL MPA ANALYSIS...

### I can run stats on the data after removing the geometry column.......but first i have to join the geospatial data together to get the mpas overlain with the sub....

#### BUT i can do it in 3s, remove the geometry, and then join the three simple dfs to run analysis, without doing too much comp. NO NEED to make a big statewide map..............at least not yet

## BUT FIRST IM GOING TO HAVE TO BREAK THIS DOWN INTO SMALLER CATEGORIES because there are some invalid polygons preventing analysis, and st_make_valid was not gonna run on a file this size. 


```{r}
# Filter to North
north_sub <- depth_zones_sub_filtered |>
  filter(pmep_region == "Pacific Northwest")

# Filter to Central
central_sub <- depth_zones_sub_filtered |>
  filter(pmep_region == "Central California")

# Filter to Southern
south_sub <- depth_zones_sub_filtered |>
  filter(pmep_region == "Southern California Bight")
```

```{r}
north_sub_mpa <- st_intersection(mpas, north_sub)
```

```{r}
central_sub_mpa <- st_intersection(mpas, central_sub)
```

```{r}
south_sub_mpa <- st_intersection(mpas, south_sub)
```

## All of the mpas and substrates were able to be combined in the separate regions....SO THE INVALID POLYGONS ARE PROBABLY IN THE REGION OVERLAP

```{r}
# Make NON GEO dfs for each region
north_sub_mpa_simple <- north_sub_mpa |> 
  st_drop_geometry()

central_sub_mpa_simple <- central_sub_mpa |> 
  st_drop_geometry()

south_sub_mpa_simple <- south_sub_mpa |> 
  st_drop_geometry()
```

```{r}
# COMBINE the simple dfs into one big simple df for all of CA depth zones!!!!!!!
california_depths_simple <- bind_rows(north_sub_mpa_simple, central_sub_mpa_simple, south_sub_mpa_simple)
```

```{r}
unique(california_depths_simple$shortname)
```

#### I am going to calculate the percentages of each MPA that are in each depth zone


```{r}

# Step 2: Make sure pmep_zone is treated as a factor (categorical)
california_depths_simple$pmep_zone_detail <- as.factor(california_depths_simple$pmep_zone_detail)

# Step 3: Calculate the percentages for each shortname and pmep_zone combination
depth_zones_mpas <- california_depths_simple |>
  # Group by shortname and pmep_zone to count occurrences
  group_by(shortname, pmep_zone_detail) |>
  summarise(count = n(), .groups = "drop") |>
  
  # Group by shortname to calculate total count for each shortname
  group_by(shortname) %>%
  mutate(percentage = count / sum(count) * 100) |>
  
  # Reshape data to have a column for each unique pmep_zone
  select(-count) |>
  pivot_wider(
    names_from = pmep_zone_detail,
    values_from = percentage) |>
    #values_fill = 0)  |> # Fill NA values with 0
  
  # Rename columns to also reflect PMEP zone code
  rename(
    "Marine Protected Area" = "shortname",
    "Zone 6: Deep Shelf or Sound (-100m to -200m), State Waters" = "Deep Shelf or Sound (-100m to -200m), State Waters" ,
    "Zone 8: Outside PMEP Scope (>-200m) or International Waters" = "Outside PMEP Scope (>-200m) or International Waters",
    #"No Data" =  "NA",
    "Zone 5: Seaward Zone (-30m to -100m), Federal" = "Seaward Zone (-30m to -100m), Federal",
    "Zone 2: Core Zone (Shoreline to -30m), State Waters" = "Core Zone (Shoreline to -30m), State Waters",
    "Zone 0: Landward Zone" = "Landward Zone",
    "Zone 4: Seaward Zone (-30m to -100m), State Waters" = "Seaward Zone (-30m to -100m), State Waters",
    "Zone 7: Deep Shelf or Sound (-100m to -200m), Federal" = "Deep Shelf or Sound (-100m to -200m), Federal",
    "Zone 3: Core Zone (Shoreline to -30m), Federal" = "Core Zone (Shoreline to -30m), Federal"
  ) |>
  # Put the columns in desired order
  select(
    "Marine Protected Area",
    "Zone 0: Landward Zone",
    "Zone 2: Core Zone (Shoreline to -30m), State Waters",
    "Zone 3: Core Zone (Shoreline to -30m), Federal",
    "Zone 4: Seaward Zone (-30m to -100m), State Waters",
    "Zone 5: Seaward Zone (-30m to -100m), Federal",
    "Zone 6: Deep Shelf or Sound (-100m to -200m), State Waters",
    "Zone 7: Deep Shelf or Sound (-100m to -200m), Federal",
    "Zone 8: Outside PMEP Scope (>-200m) or International Waters"
  )
  
```


```{r}
# Make a pivot longer table
depth_zones_long <- depth_zones_mpas %>%
  pivot_longer(
    cols = -`Marine Protected Area`, 
    names_to = "PMEP_Zone",
    values_to = "Percentage"
  )
```


```{r}
# Calculate mean percentage for each PMEP zone, ignoring NA values
total_depth_avg <- depth_zones_long %>%
  group_by(PMEP_Zone) %>%
  summarise(
    avg_percentage = mean(Percentage, na.rm = TRUE),
    n_mpas = sum(!is.na(Percentage)),  # Count of MPAs with data for this zone
    total_mpas = n_distinct(`Marine Protected Area`, na.rm = TRUE)  # Total MPAs in dataset
  ) %>%
  mutate(
    coverage = n_mpas / total_mpas * 100  # Percentage of MPAs that have this zone
  )

# Sort by average percentage...
total_depth_avg <- total_depth_avg %>%
  arrange(desc(avg_percentage))
```

```{r}
# Make a plot!!
ggplot(total_depth_avg, aes(x = reorder(PMEP_Zone, avg_percentage), y = avg_percentage)) +
  geom_col(fill = '#09EEC8') +
  coord_flip() +  # Horizontal bars
  labs(
    title = "What is the Average % Cover of any given **Depth Zone**<br>within California MPAs?",
    x = NULL,
    y = "Average Percentage Composition of Depths per Zone"
  ) +
  theme_bw() +
  theme(plot.title = element_markdown(size = 30, hjust = 1, family = "montserrat"),
        axis.text.x = element_text(family = "montserrat", size = 14),  
        axis.text.y = element_text(family = "montserrat", size = 14),  
        axis.title.x = element_text(family = "montserrat", size = 16),  
        axis.title.y = element_text(family = "montserrat", size = 16)) 
```











