---
title: "California Statewide Depth Zones Analysis"
author: "Bailey Jørgensen"
format: html
---

```{r}
# load libraries
librarian::shelf(tidyverse, janitor, sf, terra, tmap, knitr, showtext, gt, ggtext, leaflet, here)
```

## Here is what the PMEP depth zones stand for, according to the metadata:

### NOTE: The substrate data records the whole name, rather than using the # code

-   Value 0: Landward Zone
-   Value 1: Core Zone (Estuary Overlap)
-   Value 2: Core Zone (Shoreline to -30m), State Waters
-   Value 3: Core Zone (Shoreline to -30m), Federal Waters
-   Value 4: Seaward Zone (-30m to -100m), State Waters
-   Value 5: Seaward Zone (-30m to -100m), Federal Waters
-   Value 6: Deep Shelf or Channel (-100m to -200m), State Waters
-   Value 7: Deep Shelf or Channel (-100m to -200m), Federal WatersLN
-   Value 8: Outside PMEP Score (>-300m) or International Waters

## For the purposes of the PMEP depth zone analysis, I have decided to use the PMEP Zones that are recorded in the substrate data, rather than the biotic data. This is because it is expected for there to be 100% MPA coverage of the different substrate types, where as the biotic data may not contain whole coverage of MPA, given the nature of biota.

## Read in substrate, biota and MPA data

```{r}
# Read in substrate and biotic data ... Commented out, because I chose to load the RDS files into my environment prior to this
substrate <- readRDS(here("data", "substrate.rds"))
biota <- readRDS(here("data", "biota.rds")) 

# Read in MPA boundaries data
boundary.dir <- "/capstone/marinebiomaps/data/MPA_boundaries"
mpa_boundaries <- sf::st_read(file.path(boundary.dir, "California_Marine_Protected_Areas_[ds582].shp"))

# Clean and transform MPA boundaries data
mpas <- mpa_boundaries |> 
  clean_names() |> 
  select("type", "shortname", "geometry") |> 
  st_transform(mpas, crs = st_crs(substrate)) |> 
  st_make_valid()

# Filter the substrate data to just the variables I'm interested in - KEEP GEO
depth_zones_sub_filtered <- substrate |>
  select("pmep_zone_detail", "Shape", "pmep_region", "area_ha", "shape_area")
```

## OK HERE... I AM TRYING TO COMBINE THE DATA SO THAT I CAN DO FULL MPA ANALYSIS...

### I can run stats on the data after removing the geometry column.......but first i have to join the geospatial data together to get the mpas overlain with the sub....

#### BUT i can do it in 3s, remove the geometry, and then join the three simple dfs to run analysis, without doing too much comp. NO NEED to make a big statewide map..............at least not yet

## BUT FIRST IM GOING TO HAVE TO BREAK THIS DOWN INTO SMALLER CATEGORIES because there are some invalid polygons preventing analysis, and st_make_valid was not gonna run on a file this size.

```{r}
# Filter to North
north_sub <- depth_zones_sub_filtered |>
  filter(pmep_region == "Pacific Northwest")

# Filter to Central
central_sub <- depth_zones_sub_filtered |>
  filter(pmep_region == "Central California")

# Filter to Southern
south_sub <- depth_zones_sub_filtered |>
  filter(pmep_region == "Southern California Bight")
```

```{r}
north_sub_mpa <- st_intersection(mpas, north_sub)
```

```{r}
central_sub_mpa <- st_intersection(mpas, central_sub)
```

```{r}
south_sub_mpa <- st_intersection(mpas, south_sub)
```

## All of the mpas and substrates were able to be combined in the separate regions....SO THE INVALID POLYGONS ARE PROBABLY IN THE REGION OVERLAP

```{r}
# Make NON GEO dfs for each region
north_sub_mpa_simple <- north_sub_mpa |> 
  st_drop_geometry()

central_sub_mpa_simple <- central_sub_mpa |> 
  st_drop_geometry()

south_sub_mpa_simple <- south_sub_mpa |> 
  st_drop_geometry()
```

```{r}
# COMBINE the simple dfs into one big simple df for all of CA depth zones!!!!!!!
california_depths_simple <- bind_rows(north_sub_mpa_simple, central_sub_mpa_simple, south_sub_mpa_simple)
```

```{r}
unique(california_depths_simple$shortname)
```

#### I am going to calculate the percentages of each MPA that are in each depth zone

```{r}

# Step 2: Make sure pmep_zone is treated as a factor (categorical)
california_depths_simple$pmep_zone_detail <- as.factor(california_depths_simple$pmep_zone_detail)

# Step 3: Calculate the percentages for each shortname and pmep_zone combination
depth_zones_mpas <- california_depths_simple |>
  # Group by shortname and pmep_zone to count occurrences
  group_by(shortname, pmep_zone_detail) |>
  summarise(count = n(), .groups = "drop") |>
  
  # Group by shortname to calculate total count for each shortname
  group_by(shortname) %>%
  mutate(percentage = count / sum(count) * 100) |>
  
  # Reshape data to have a column for each unique pmep_zone
  select(-count) |>
  pivot_wider(
    names_from = pmep_zone_detail,
    values_from = percentage) |>
    #values_fill = 0)  |> # Fill NA values with 0
  
  # Rename columns to also reflect PMEP zone code
  rename(
    "Marine Protected Area" = "shortname",
    "Zone 6" = "Deep Shelf or Sound (-100m to -200m), State Waters" ,
    "Zone 8" = "Outside PMEP Scope (>-200m) or International Waters",
    #"No Data" =  "NA",
    "Zone 5" = "Seaward Zone (-30m to -100m), Federal",
    "Zone 2" = "Core Zone (Shoreline to -30m), State Waters",
    "Zone 0" = "Landward Zone",
    "Zone 4" = "Seaward Zone (-30m to -100m), State Waters",
    "Zone 7" = "Deep Shelf or Sound (-100m to -200m), Federal",
    "Zone 3" = "Core Zone (Shoreline to -30m), Federal"
  ) |>
  # Put the columns in desired order
  select(
    "Marine Protected Area",
    "Zone 0",
    "Zone 2",
    "Zone 3",
    "Zone 4",
    "Zone 5",
    "Zone 6",
    "Zone 7",
    "Zone 8"
  )
  
```

# This does the same thing, but makes it a long table instead of wide

```{r}
# Make a pivot longer table
depth_zones_long <- depth_zones_mpas %>%
  pivot_longer(
    cols = -`Marine Protected Area`, 
    names_to = "PMEP_Zone",
    values_to = "Percentage"
  )
```













# More Visuals for Report: 

```{r}
# Leaflet requrires a CRS of WGS84
mpa_boundaries_WGS84 <- mpa_boundaries |> 
  sf::st_transform('+proj=latlong +datum=WGS84')

# Lets just make a map of all of the CA MPAs
leaflet() |>
  addProviderTiles(providers$CartoDB.Positron) |>
  setView(lng = -119.6989,
          lat = 34.4203,
          zoom = 8) |>
  addMiniMap(toggleDisplay = TRUE,
             minimized = FALSE) |>
  addPolygons(data = mpa_boundaries_WGS84,
              color = "darkgoldenrod",
              weight = 3,
              popup = paste0("MPA Name:  ", mpa_boundaries_WGS84$FULLNAME, "<br>",
                             "Total Square Miles:  ", mpa_boundaries_WGS84$Area_sq_mi))
```



